---
title: "Democracy-in-action"
description: I don't know what I'm more excited about - new election results to pore over, or my first follow-up post! In 2016 I looked at voter turnout in the New Zealand local body elections. Now three years later, almost to the day, let's have a look again and see how things have turned out this year. Is it Democracy In Action or Democracy Inaction?
author: "David Friggens"
date: 2019-10-13
preview: images/preview.png
categories:
  - all
  - R
  - elections
  - New Zealand
  - rpkg plotly
  - dumbbell plots
output:
  distill::distill_article:
    self_contained: false
---

Three years ago [I looked at voter turnout in the 2016 local body elections](../2016-10-whither-local-democracy)
in New Zealand. There was a lot of concern about the low voter turnout, though
overall the turnout was slightly higher than 2013. This was driven by the
"metro" councils, which countered the decline in the "provincial" and "rural"
councils.

Fellow kiwis will likely have heard similar rumblings about dropping voter
turnout again, so let's see what actually happened.
Luckily I was reminded from my previous post that Local Government NZ
[collates all the data](https://www.lgnz.co.nz/vote2019/voters/preliminary-voter-turnout-2019/),
which saved me the nightmare of visiting every individual council's website.
You can go to their page to see some Excel column charts, or you can stay
here for a repeat of the dumbbell plots.

<aside>
Yes, it's 2019 and local council websites are still an eclectic mix of
volume, frequency and (especially) file types when it comes to voting data.
</aside>


```{r setup}
library(readr)
library(dplyr)
library(tidyr)
library(plotly)

# last time I scraped it with rvest
# this time I manually appended the new data
summary_turnout <- read_csv("data/summary_turnout.csv")
council_turnout <- read_csv("data/council_turnout.csv")
```


## High-level summaries

After a big drop between 2010 and 2013 and a smaller movement in 2016, there
appears to be another big drop in turnout across the board.
The national figures apparently suggest an overall increase, but this must
be a mistake. These are preliminary figures, and I'll update when final numbers
are confirmed.


```{r turnout_summary, fig.height=2}
nto <- 
  summary_turnout %>% 
  spread(Year, Turnout) %>% 
  mutate(min_to = pmin(`2010`, `2013`, `2016`, `2019`),
         max_to = pmax(`2010`, `2013`, `2016`, `2019`))
plot_ly(nto,
        color = I("gray80")) %>% 
  add_segments(x = ~min_to, xend = ~max_to, y = ~Group, yend = ~Group, showlegend = FALSE) %>% 
  add_markers(x = ~`2010`, y = ~Group, name = "2010", color = I("red"),
              hoverinfo = "text",
              text = ~paste0(`2010`, "% turnout ", if_else(Group == "National", "across New Zealand", paste0("in ", Group, " councils")), " in 2010")) %>% 
  add_markers(x = ~`2013`, y = ~Group, name = "2013", color = I("yellow"),
              hoverinfo = "text",
              text = ~paste0(`2013`, "% turnout ", if_else(Group == "National", "across New Zealand", paste0("in ", Group, " councils")), " in 2013")) %>% 
  add_markers(x = ~`2016`, y = ~Group, name = "2016", color = I("blue"),
              hoverinfo = "text",
              text = ~paste0(`2016`, "% turnout ", if_else(Group == "National", "across New Zealand", paste0("in ", Group, " councils")), " in 2016")) %>% 
  add_markers(x = ~`2019`, y = ~Group, name = "2019", color = I("black"),
              hoverinfo = "text",
              text = ~paste0(`2019`, "% turnout ", if_else(Group == "National", "across New Zealand", paste0("in ", Group, " councils")), " in 2019")) %>% 
  layout(
    title = "Local Election Turnout in New Zealand",
    xaxis = list(title = "Voter Turnout (Percent)"),
    yaxis = list(title = "")
  ) %>% 
  config(displayModeBar = FALSE)
```

<aside>
The four-value dumbbells hopefully aren't too cluttered. I was going to drop
2010 off to make it cleaner, but I think it's better to keep it to make clearer
the story of an ongoing drop.
</aside>

Whilst it's positive to see that rural and provincial districts have good
turnout relative to the metro areas, they only make up a third of the population.
To put this in perspective, let's turn these percentages into people.

```{r pc2ppl}
population <- read_csv("data/population_2018_age15plus.csv", skip = 1)
council_turnout %>%
  filter(Year == 2019) %>%
  inner_join(population, by = c(Council = "Area")) %>%
  mutate(Voted = Population * (Turnout / 100),
         DidntVote = Population - Voted) %>%
  group_by(Type) %>%
  summarise(Voted = round(sum(Voted), 0),
            DidntVote = round(sum(DidntVote), 0)) %>%
  ungroup() %>% 
  inner_join(summary_turnout %>% filter(Year == 2019),
             by = c(Type = "Group")) %>% 
  plot_ly(x = ~Type,
          y = ~Voted,
          text = ~paste0("In ", Type, " councils turnout was ", Turnout, "%.<br>",
                         "Approximately ", prettyNum(Voted %/% 1000 * 1000, big.mark = ","), " people voted."),
          type = 'bar',
          name = 'People that voted',
          hoverinfo = "text") %>% 
  add_trace(y = ~DidntVote,
            text = ~paste0("In ", Type, " councils turnout was ", Turnout, "%.<br>",
                           "Approximately ", prettyNum(DidntVote %/% 1000 * 1000, big.mark = ","), " people didn't vote."),
            name = "People that didn't vote",
            hoverinfo = "text") %>% 
  layout(
    title = "Local Election Turnout in New Zealand, 2019",
    xaxis = list(title = ""),
    yaxis = list(title = "Population"),
    barmode = 'stack',
    legend = list(x = 0.7, y = 0.95)
  ) %>% 
  config(displayModeBar = FALSE)
```

<aside>
NB I've used territorial authority populations for age 15+, rather than the
number of voters (who must be at least 18) so the numbers aren't precise, but
the relative comparison should be accurate.
</aside>



## Individual councils

We can look at individual councils for a finer breakdown.
Here are dumbells for the last three elections grouped by metro, provincial
and rural, then ordered by the 2019 turnout.

```{r turnout_dumbbell0, fig.height=12}
cto <- 
  council_turnout %>% 
  spread(Year, Turnout) %>% 
  mutate(min_to = pmin(`2013`, `2016`, `2019`),
         max_to = pmax(`2013`, `2016`, `2019`))
cto$Council <- 
  factor(cto$Council,
         levels = cto$Council[order(cto$Type, cto$`2019`, method = "radix", decreasing = c(TRUE, FALSE))])
plot_ly(cto,
        color = I("gray80")) %>% 
  add_segments(x = ~min_to, xend = ~max_to, y = ~Council, yend = ~Council, showlegend = FALSE) %>% 
  add_markers(x = ~`2013`, y = ~Council, name = "2013", color = I("yellow"),
              hoverinfo = "text",
              text = ~paste0(`2013`, "% turnout in ", Council, " in 2013")) %>% 
  add_markers(x = ~`2016`, y = ~Council, name = "2016", color = I("blue"),
              hoverinfo = "text",
              text = ~paste0(`2016`, "% turnout in ", Council, " in 2016")) %>% 
  add_markers(x = ~`2019`, y = ~Council, name = "2019", color = I("black"),
              hoverinfo = "text",
              text = ~paste0(`2019`, "% turnout in ", Council, " in 2019")) %>% 
  layout(
    title = "Local Election Turnout in New Zealand",
    xaxis = list(title = "Voter Turnout (Percent)"),
    yaxis = list(title = ""),
    margin = list(l = 180)
  ) %>% 
  config(displayModeBar = FALSE)
```

<aside>
2019 turnout figures haven't been released yet for Porirua, New Plymouth or
Clutha. I'll update these in the coming days.
</aside>

As expected there is a lot of variation, with big jumps both up and down over 2016.
Most areas record a drop though, with the largest increases generally being in small districts.
Only two metro areas had (very small) increases: Christchurch and [Lower] Hutt.
The biggest metro drops were in Auckland and Wellington --- two of the three
biggest areas.


## Data {.appendix}

Turnout data [by area type](data/summary_turnout.csv) and
[by council](data/council_turnout.csv), sourced
[from LGNZ](https://www.lgnz.co.nz/vote2019/voters/preliminary-voter-turnout-2019/) on 13 October 2019.
The 2016 numbers may differ from my previous post as I never updated the
preliminary figures when final ones were announced.

2018 [population data](data/population_2018_age15plus.csv)
for council areas sourced from [Stats NZ](https://nzdotstats.stats.govt.nz).


