---
title: "Anyone Who Had a Heart"
date: "29 January 2017"
output:
  html_document:
    df_print: kable
---

```{r opts, echo = FALSE}
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(
  fig.path = "images/"
)
```

One of my goals for the year is to blog at least once a month - and time
is nearly up for January! I've got a couple of things I've started on,
notably my "quantified self" data, though I'm still [letting the data
ramble](https://flowingdata.com/2017/01/24/one-dataset-visualized-25-ways/).
Inspired by Maëlle Salmon's [Catan visualisation](http://www.masalmon.eu/2017/01/27/catan/)
I thought I'd finish up a visualisation of Hearts that I poked at last
weekend. 

I've played a fair bit of Hearts over the years, and occasionally wondered
about visualising the results. I played two games last weekend with my
parents; the following morning, left at home sick whilst everyone else
went to the National Aquarium, I thought I'd give it a go.

I'm not convinced they particularly add anything, but sometimes you have
to try in order to find out.

# Hearts

If you don't know [Hearts](https://en.wikipedia.org/wiki/Hearts),
it's fairly straightforward. Four players
are dealt all the cards and play 13 hands, with the highest card in
the suit led winning. Points are for certain cards won, not for the
number of hands. Each heart is 1 point (hence the name) and the Queen
of Spades is 13 points. The catch is that points are bad --- the first
to 100 points loses, and the player with the lowest score is the winner.

So there are 26 points to be distributed each round, which means the
game has to take at least 4 rounds, and (I think) can't last any more
than 16 rounds. The other catch is that if one player gets all 26 points
then they score 0 and the other three score 26. This is called "shooting
the moon" and is harder than it sounds.^[I managed it in one of these
games, but it's the first time I've been successful in ages.]

OK, so let's load the data. I saved the points per round, and we can
calculate the scores as we go, taking into account possibly shooting
the moon.

```{r load_data}
library(readr)
library(dplyr)
points1 <- 
  read_csv("data/hearts-g1.csv") %>% 
  mutate(round = row_number())
points2 <- 
  read_csv("data/hearts-g2.csv") %>% 
  mutate(round = row_number())
shoot_the_moon <- 
  function(df) {
    df %>% 
      # Lazily hard-code column names
      mutate(zeros = (B==0) + (O==0) + (N==0) + (D==0)) %>%
      mutate(B = ifelse(zeros == 3, ifelse(B == 0, 26, 0), B),
             O = ifelse(zeros == 3, ifelse(O == 0, 26, 0), O),
             N = ifelse(zeros == 3, ifelse(N == 0, 26, 0), N),
             D = ifelse(zeros == 3, ifelse(D == 0, 26, 0), D)) %>% 
      select(-zeros)
  }
score1 <- 
  points1 %>% 
  shoot_the_moon() %>% 
  mutate_at(1:4, cumsum)
score2 <- 
  points2 %>% 
  shoot_the_moon() %>% 
  mutate_at(1:4, cumsum)
```

The standard way to look at the results is just to tabulate the scores
at each round. The first game took 8 rounds, and I managed to win by
a comfortable margin. I shot the moon successfully in round 5, and
tried again less successfully in round 7. 
The second game was shorter, due to one player lucking out with the
queen of spades nearly every round.
<div class = "row">
<div class = "col-md-6">
```{r table1}
score1 %>% select(-round)
```
</div><div class = "col-md-6">
```{r table2}
score2 %>% select(-round)
```
</div></div>

If we want to graph anything, we're going to need
[tidy data](ftp://cran.r-project.org/pub/R/web/packages/tidyr/vignettes/tidy-data.html).

```{r tidy}
library(tidyr)
library(magrittr)
points1 %<>% gather(player, points, -round)
points2 %<>% gather(player, points, -round)
score1 %<>% gather(player, score, -round)
score2 %<>% gather(player, score, -round)
```


# Cumulative score {.tabset}

The most obvious first step to me is to graph the scores as they accumulate.
Normally for this sort of thing I would use a line graph, but with the
discrete jumps of the rounds I wondered if a step graph would work better.
Looking at them, I'm still on the fence, so I've included both --- see the
tabs below.

Either way, I think it makes sense to reverse the y axis, given that points
are effectively negative. And having a line to indicate the end-game
condition of 100 points makes it clearer.

I was tempted to highlight the player who got the Queen of Spades in each
round, but since that's half the points on offer I think it's reasonably
obvious in each case.


```{r ggplot}
library(ggplot2)
library(ggthemes)
```


## Game 1, Steps
```{r 2017-01-steps-1}
ggplot(score1, aes(x = round, y = score, group = player, color = player)) +
  geom_step(size = 1) + geom_hline(yintercept = 100, color = "black") + 
  scale_x_continuous(breaks=1:8) + scale_y_reverse(breaks = seq(0, 110, 10))
```

## Game 1, Lines
```{r 2017-01-lines-1}
ggplot(score1, aes(x = round, y = score, group = player, color = player)) +
  geom_line() + geom_point() + geom_hline(yintercept = 100, color = "black") +
  scale_x_continuous(breaks=1:8) + scale_y_reverse(breaks = seq(0, 110, 10))
```

## Game 2, Steps
```{r 2017-01-steps-2}
ggplot(score2, aes(x = round, y = score, group = player, color = player)) +
  geom_step(size = 1) + geom_hline(yintercept = 100, color = "black") +
  scale_x_continuous(breaks=1:8) + scale_y_reverse(breaks = seq(0, 110, 10))
```

## Game 2, Lines
```{r 2017-01-lines-2}
ggplot(score2, aes(x = round, y = score, group = player, color = player)) +
  geom_line() + geom_point() + geom_hline(yintercept = 100, color = "black") +
  scale_x_continuous(breaks=1:8) + scale_y_reverse(breaks = seq(0, 110, 10))
```

# Rank {.tabset}

Hiding inside the plots of the scores are the stories of the ranking
leaderboard. We can look at the positions explicitly, to more easily
see when/if they swap around.


```{r}
graph_by_rank <- 
  function(df) {
    df %>% 
      group_by(round) %>% 
      mutate(rank = min_rank(score)) %>% 
      ungroup() %>% 
      ggplot(aes(x = round, y = rank, group = player, colour = player)) +
      geom_line(size = 2) + 
      scale_x_continuous(breaks = 1:8) +
      scale_y_reverse() +
      theme_tufte()
  }
```


## Game 1
```{r 2017-01-rank-1}
graph_by_rank(score1)
```

## Game 2
```{r 2017-01-rank-2}
graph_by_rank(score2)
```


# Points {.tabset}

Given that there are the same number of points to be shared each round,
we can plot the share of points per player.
It gives you a clearer look at the results per round, and the variation
between them, than the cumulative plots above.
You get less of an insight into the game overall, but (at least for
these two) it's fairly clear who came last.

## Game 1
```{r 2017-01-points-1}
ggplot(points1, aes(x = round, y = points, group = player, fill = player)) +
  geom_col() + theme_tufte() +
  scale_x_continuous(breaks = 1:8) + scale_y_continuous(breaks = seq(0, 25, 5))
```

## Game 2
```{r 2017-01-points-2}
ggplot(points2, aes(x = round, y = points, group = player, fill = player)) +
  geom_col() + theme_tufte() +
  scale_x_continuous(breaks = 1:8) + scale_y_continuous(breaks = seq(0, 25, 5))
```

# Reflection

I'll be honest --- before I started I thought these plots would be more
interesting than they turned out. They're of some interest to the
participants, but I don't think they add terribly much on top of the
plain tables.

In part, I think this is because of the small number of rounds. For a
more closely fought game of Hearts, or for a similar but longer card game,
I think the graphs might be a bit more interesting.

I'm going to continue playing Hearts, but I think this is the last time
I'll bother trying to visualise the results.
